---
title: "HW5"
author: "C. Nicholson"
date: "9/26/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Using `dplyr` to manipulate data  
##### Preliminaries  
``` {r, echo = T, message=FALSE}
require(dplyr)
```  

##### Examine the structure of the `iris` dataset. How many observations and variables are in the dataset?  
``` {r}
glimpse(iris)  
```  

##### Create a new data frame `iris1` that contains only the species virginica and versicolor with sepal lengths longer than 6 cm and sepal widths longer than 2.5 cm. How many observations and variables are in the dataset?  
``` {r}

iris1 <- filter(iris, 
       Species %in% c("virginica", "versicolor"), 
       Sepal.Length > 6,
       Sepal.Width > 2.5)
glimpse(iris1)  

```  
##### Now, create a `iris2` data frame from `iris1` that contains only the columns for Species, `Sepal.Length`, and `Sepal.Width`. How many observations and variables are in the dataset?  


``` {r}

iris2 <- select(iris1, 
                c(Species, Sepal.Length, Sepal.Width))
glimpse(iris2) 

```  

##### Create an `iris3` data frame from `iris2` that orders the observations from largest to smallest sepal length. Show the first 6 rows of this dataset.  


``` {r}

iris3 <- arrange(iris2, -desc(Sepal.Length))

head(iris3) 

```  

##### Create an `iris4` data frame from `iris3` that creates a column with a sepal area (length * width) value for each observation. How many observations and variables are in the dataset?  


``` {r}

iris4 <- mutate(iris3, Sepal.Area = Sepal.Length * Sepal.Width)

glimpse(iris4)

```  

##### Create `iris5` that calculates the average sepal length, the average sepal width, and the sample size of the entire `iris4` data frame and print `iris5`.  


``` {r}

iris5 <- summarize(iris4, 
                   meanLength = mean(Sepal.Length), 
                   meanWidth = mean(Sepal.Width), 
                   meanArea = mean(Sepal.Area), 
                   N = n()) 
print(iris5)


```  

##### Finally, create `iris6` that calculates the average sepal length, the average sepal width, and the sample size for each species of in the `iris4` data frame and print `iris6`.  


``` {r}

iris6 <- iris4 %>%
  group_by(Species) %>%
  summarise(meanLength = mean(Sepal.Length), 
                   meanWidth = mean(Sepal.Width), 
                   meanArea = mean(Sepal.Area), 
                   N = n())
print(iris6)

```

##### A more efficient way to do this is to use the pipe operator `%>%` from the tidyr package. See if you can rework all of your previous statements into an extended piping operation that uses iris as the input and generates `iris6` as the output.  


``` {r}

iris6 <- iris %>%
  filter(Species %in% c("virginica", "versicolor"), 
       Sepal.Length > 6,
       Sepal.Width > 2.5) %>%
  mutate(Sepal.Area = Sepal.Length * Sepal.Width) %>%
  group_by(Species) %>%
  summarise(meanLength = mean(Sepal.Length), 
                   meanWidth = mean(Sepal.Width), 
                   meanArea = mean(Sepal.Area), 
                   N = n())
print(iris6)

```